---
title: "Complete Workflow: From Data to Decision"
author: "Deniz Akdemir"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete Workflow: From Data to Decision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
eval_surv_effect <- getRversion() >= "4.0.0" && requireNamespace("survival", quietly = TRUE)

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5
)

if (!exists("deparse1", envir = baseenv())) {
  deparse1 <- function(expr, collapse = " ", width.cutoff = 500L, ...) {
    paste(deparse(expr, width.cutoff, ...), collapse = collapse)
  }
}

fmt_num <- function(x, digits = 3) {
  formatC(x, digits = digits, format = "f")
}
```

## Overview

This vignette demonstrates the complete **causaldef** workflow, from data 
specification through to policy decision-making. We show how deficiency 
theory translates abstract statistical concepts into actionable clinical 
insights.

The workflow consists of four stages:

1. **Specify** → Define the causal problem
2. **Estimate** → Compute deficiency for different adjustment strategies  
3. **Diagnose** → Validate assumptions using negative controls and sensitivity analysis
4. **Decide** → Compute policy regret bounds and make informed decisions

---

## Part 1: Gene Perturbation Study (Continuous Outcome)

We begin with the `gene_perturbation` dataset, which simulates a CRISPR 
knockout experiment. This illustrates the core workflow for continuous outcomes.

### 1.1 Data Description

```{r load-data-gene}
library(causaldef)
library(ggplot2)

data(gene_perturbation)
str(gene_perturbation)
```

**Variables:**

- `knockout_status`: Treatment (Control vs. Knockout)
- `target_expression`: Primary outcome (gene expression level)
- `housekeeping_gene`: Negative control outcome (shouldn't be affected by knockout)
- `batch`, `library_size`: Technical confounders

**Causal Structure:**

```
      [Batch, Library Size]
             |
             v
   [Knockout] -----> [Target Expression]
             \
              \---X--> [Housekeeping Gene]  (no causal effect)
```

The housekeeping gene is affected by the same technical variations but NOT by 
the knockout, making it an ideal negative control.

### 1.2 Step 1: Specification

```{r spec-gene}
spec_gene <- causal_spec(
    data = gene_perturbation,
    treatment = "knockout_status",
    outcome = "target_expression",
    covariates = c("batch", "library_size"),
    negative_control = "housekeeping_gene",
    estimand = "ATE",
    outcome_type = "continuous"
)

print(spec_gene)
```

### 1.3 Step 2: Deficiency Estimation

We compare three adjustment strategies:

1. **Unadjusted**: Ignores technical confounders
2. **IPTW**: Reweights samples to balance batch and library size
3. **AIPW**: Augmented IPTW (doubly robust)

```{r estim-gene}
deficiency_gene <- estimate_deficiency(
    spec_gene,
    methods = c("unadjusted", "iptw", "aipw"),
    n_boot = 100 # Use more for production (e.g., 1000)
)

print(deficiency_gene)
```

**Interpretation:**

In this run, the unadjusted PS-TV proxy is about `r fmt_num(deficiency_gene$estimates["unadjusted"])`, while IPTW and AIPW reduce it to about `r fmt_num(deficiency_gene$estimates["iptw"])`. That is a substantial reduction in the observational-to-interventional gap, but it should still be interpreted through the downstream regret bounds rather than as literal proof of randomized equivalence.

```{r plot-gene, fig.height=4}
plot(deficiency_gene, type = "bar")
```

### 1.4 Step 3: Diagnose with Negative Control

The negative control diagnostic tests whether our adjustment removes ALL 
confounding, not just the measured confounders.

```{r nc-gene}
set.seed(123)
nc_test <- nc_diagnostic(
    spec_gene,
    method = "iptw",
    alpha = 0.05,
    n_boot = 100
)

print(nc_test)
```

**Decision Logic:**

| Result | Interpretation | Action |
|--------|----------------|--------|
| `falsified = FALSE` | The screen does not find strong evidence against the current adjustment, but this is not proof that all confounding is gone | Proceed with explicit uncertainty qualification |
| `falsified = TRUE` | Residual confounding is detected by the screen | Add covariates, reconsider the design, or report the limitation |

### 1.5 Step 4: Policy Decision

Suppose we're deciding whether to pursue this gene target for drug development. 
The utility is measured on a scale where:
- 0 = no promise (no effect on expression)
- 10 = maximum promise (strong effect)

```{r policy-gene}
bounds_gene <- policy_regret_bound(
    deficiency_gene,
    utility_range = c(0, 10)
)

print(bounds_gene)
```

**Regret Bounds:**

`policy_regret_bound()` reports:

- **Transfer penalty** \(M\cdot\delta\): additive worst-case regret inflation term, and
- **Minimax safety floor** \((M/2)\cdot\delta\): irreducible worst-case regret when \(\delta>0\).

**Decision Rule:**
- If `transfer_penalty` is small relative to the practical stakes of the decision → the observational evidence may be usable with caveats
- If `transfer_penalty` is large relative to those stakes → seek more evidence or narrow the decision scope

### 1.6 Effect Estimation

Finally, we estimate the causal effect using the best-performing method:

```{r effect-gene}
effect_gene <- estimate_effect(
    deficiency_gene,
    target_method = "iptw"
)

print(effect_gene)
```

**Complete Report:**

```{r report-gene, results='asis'}
conclusion_gene <- if (nc_test$falsified) {
    "Residual confounding was detected by the negative control screen. Treat the effect estimate as exploratory."
} else if (bounds_gene$transfer_penalty < 0.5) {
    "Adjustment reduced the proxy gap substantially and the negative control did not falsify the analysis, but residual uncertainty remains."
} else {
    "Adjustment improved balance, but the remaining proxy gap is still material on the chosen utility scale."
}

cat(sprintf(
    "
## Gene Perturbation Analysis Report

**Treatment Effect (IPTW-adjusted):** %.2f log2 expression units
**Deficiency (δ):** %.3f
**Negative Control Screen:** %s (p = %.3f; delta_NC = %.3f)
**Transfer Penalty:** %.3f on [0, 10] scale
**Minimax Safety Floor:** %.3f on [0, 10] scale

**Conclusion:** %s
",
    effect_gene$estimate,
    deficiency_gene$estimates["iptw"],
    ifelse(nc_test$falsified, "Falsified", "Not falsified"),
    nc_test$p_value,
    nc_test$delta_nc,
    bounds_gene$transfer_penalty,
    bounds_gene$minimax_floor,
    conclusion_gene
))
```

---

## Part 2: Hematopoietic Cell Transplantation (Survival Outcome)

Next, we analyze the `hct_outcomes` dataset, which mimics a retrospective 
registry study comparing conditioning regimens in HCT.

### 2.1 Data Description

```{r load-data-hct}
data(hct_outcomes)
str(hct_outcomes)

# Summarize key variables
summary(hct_outcomes[, c("age", "kps", "time_to_event")])
table(hct_outcomes$conditioning_intensity, hct_outcomes$event_status)
```

**Clinical Context:**

- **Myeloablative conditioning**: High-intensity chemotherapy (younger, healthier patients)
- **Reduced-intensity conditioning**: Lower dose (older, sicker patients)
- **Outcome in this vignette**: 24-month restricted mean survival using death as the binary event

The key challenge is **confounding by indication**: doctors assign treatment 
based on patient status, making naive comparisons biased.

### 2.2 Step 1: Survival Specification

The deficiency, sensitivity, and regret-bound calculations in this section run on all supported runtimes. The final RMST effect-estimation chunk requires a compatible `survival` runtime, which in the current support matrix means `R >= 4.0`.

```{r spec-hct}
# Create binary event indicator for death
hct_outcomes$event_death <- as.integer(as.character(hct_outcomes$event_status) == "Death")

spec_hct <- causal_spec_survival(
    data = hct_outcomes,
    treatment = "conditioning_intensity",
    time = "time_to_event",
    event = "event_death",
    covariates = c("age", "disease_status", "kps", "donor_type"),
    estimand = "RMST",
    horizon = 24 # 24-month restricted mean survival time
)

print(spec_hct)
```

For a full competing-risks analysis of relapse versus death, prefer `causal_spec_competing()` together with `estimate_deficiency_competing()`. Here we keep the workflow aligned with the current reproducible death-endpoint survival interface.

### 2.3 Step 2: Deficiency Estimation

```{r estim-hct}
deficiency_hct <- estimate_deficiency(
    spec_hct,
    methods = c("unadjusted", "iptw"),
    n_boot = 50 # Use more for production
)

print(deficiency_hct)
```

**Clinical Interpretation:**

The deficiency tells us how much our observational evidence differs from 
what an RCT would provide. In this run, the unadjusted proxy is about `r fmt_num(deficiency_hct$estimates["unadjusted"])`, while IPTW reduces it to about `r fmt_num(deficiency_hct$estimates["iptw"])`. That is a meaningful improvement, but the remaining gap is still large enough to matter on a 24-month clinical utility scale.

### 2.4 Step 3: Confounding Frontier

Beyond point estimates, we can map a *sensitivity analysis* showing how 
deficiency varies with hypothetical unmeasured confounding:

```{r frontier-hct, fig.height=5}
frontier <- confounding_frontier(
    spec_hct,
    alpha_range = c(-2, 2), # Confounding path: U → Treatment
    gamma_range = c(-2, 2), # Confounding path: U → Outcome
    grid_size = 30
)

print(frontier)
plot(frontier)
```

**Reading the Frontier Map:**

- **Center** (α = 0 or γ = 0): No unmeasured confounding → δ = 0
- **Corners**: Strong confounding on both paths → high δ
- **Observed covariates** (dots): Benchmark strengths of measured confounders

If an unmeasured confounder would need extreme strength (beyond observed 
benchmarks) to substantially increase δ, conclusions are robust.

### 2.5 Step 4: Policy Regret and RMST Effect

```{r policy-hct}
# Utility = months of survival (horizon = 24)
bounds_hct <- policy_regret_bound(
    deficiency_hct,
    utility_range = c(0, 24)
)

print(bounds_hct)
```

**Clinical Regret Bounds:**

In the current example, the IPTW proxy implies a transfer penalty of about `r fmt_num(bounds_hct$transfer_penalty, 2)` months and a minimax safety floor of about `r fmt_num(bounds_hct$minimax_floor, 2)` months on the 0--24 month utility scale. That means the residual observational-to-interventional gap is materially improved relative to the unadjusted analysis, but still not negligible for treatment decisions.

```{r effect-hct, eval = eval_surv_effect}
# Estimate RMST difference
effect_hct <- estimate_effect(
    deficiency_hct,
    target_method = "iptw",
    contrast = c("Myeloablative", "Reduced")
)

print(effect_hct)
```

```{r effect-hct-note, results='asis', eval = !eval_surv_effect}
cat("Effect-estimation chunks are skipped on runtimes without the required survival support. The deficiency, frontier, and regret-bound calculations above still provide the main observational-versus-interventional diagnostics for this example.")
```

### 2.6 Complete Decision Framework

```{r decision-hct, results='asis', eval = eval_surv_effect}
delta_iptw <- deficiency_hct$estimates["iptw"]
transfer_penalty <- bounds_hct$transfer_penalty
minimax_floor <- bounds_hct$minimax_floor
rmst_diff <- effect_hct$estimate

# Decision logic
if (transfer_penalty < 1) {
    evidence_quality <- "LOW residual decision risk"
} else if (transfer_penalty < 2) {
    evidence_quality <- "MODERATE residual decision risk"
} else {
    evidence_quality <- "HIGH residual decision risk"
}

# Benefit-to-risk ratio
if (!is.na(rmst_diff) && !is.na(transfer_penalty) && transfer_penalty > 0) {
    benefit_to_risk <- abs(rmst_diff) / transfer_penalty
    recommendation <- ifelse(benefit_to_risk > 2,
        "The estimated effect exceeds the transfer penalty, but the recommendation should still be qualified by residual confounding risk.",
        "The estimated effect is not clearly separated from the transfer penalty; avoid strong recommendations from observational evidence alone."
    )
} else {
    benefit_to_risk <- NA
    recommendation <- "Unable to calculate benefit-to-risk ratio"
}

cat(sprintf(
    "
## HCT Treatment Decision Report

**RMST Difference (IPTW):** %.2f months (%s favored)
**Deficiency:** %.3f
**Residual Decision Risk:** %s
**Transfer Penalty:** %.2f months
**Minimax Safety Floor:** %.2f months

**Benefit-to-Risk Ratio:** %.1f:1
**Recommendation:** %s

### Clinical Translation

The observational evidence suggests %s conditioning provides approximately
%.1f months difference in restricted mean survival within the first 24 months.

However, the transfer penalty is %.1f months and the minimax safety floor is %.1f months
on the 0--24 month utility scale. This is a decision-relevant amount of residual uncertainty, so clinicians should weigh it against individual patient factors and avoid over-interpreting the observational comparison.
",
    abs(rmst_diff),
    ifelse(rmst_diff > 0, "Myeloablative", "Reduced"),
    delta_iptw,
    evidence_quality,
    transfer_penalty,
    minimax_floor,
    benefit_to_risk,
    recommendation,
    ifelse(rmst_diff > 0, "myeloablative", "reduced-intensity"),
    abs(rmst_diff),
    transfer_penalty,
    minimax_floor
))
```

---

## Part 3: Comparative Analysis Across Studies

### 3.1 When Is Observational Evidence Sufficient?

| Study | IPTW proxy | Additional diagnostic | Transfer penalty | Interpretation |
|-------|------------|---------------------|------------------|----------------|
| Gene Perturbation | `r fmt_num(deficiency_gene$estimates["iptw"])` | NC not falsified; `r fmt_num(nc_test$delta_nc)` observable residual-association proxy | `r fmt_num(bounds_gene$transfer_penalty, 2)` on [0, 10] | Residual uncertainty appears modest but nonzero |
| HCT Death-Endpoint RMST | `r fmt_num(deficiency_hct$estimates["iptw"])` | Confounding frontier recommended; no negative control in this example | `r fmt_num(bounds_hct$transfer_penalty, 2)` months | Residual uncertainty remains material on the 24-month scale |

### 3.2 General Workflow Summary

```
┌─────────────────────────────────────────────────────────────────┐
│  SPECIFY: causal_spec() / causal_spec_survival()                │
│           ↓ Define treatment, outcome, covariates, NC           │
├─────────────────────────────────────────────────────────────────┤
│  ESTIMATE: estimate_deficiency()                                │
│            ↓ Compare unadjusted, IPTW, AIPW, TMLE, etc.        │
│            ↓ Select method with lowest δ                        │
├─────────────────────────────────────────────────────────────────┤
│  DIAGNOSE: nc_diagnostic() + confounding_frontier()             │
│            ↓ Test whether assumptions are falsified             │
│            ↓ Map sensitivity to unmeasured confounding          │
├─────────────────────────────────────────────────────────────────┤
│  DECIDE: policy_regret_bound() + estimate_effect()              │
│          ↓ Compute transfer penalty / minimax floor             │
│          ↓ Report effect with uncertainty qualification         │
└─────────────────────────────────────────────────────────────────┘
```

---

## References

1. Akdemir, D. (2026). Constraints on Causal Inference as Experiment Comparison: 
   A Framework for Identification, Transportability, and Policy Learning. 
   DOI: 10.5281/zenodo.18367347

2. Le Cam, L., & Yang, G. L. (2000). Asymptotics in Statistics: Some Basic 
   Concepts. Springer.

3. VanderWeele, T. J., & Ding, P. (2017). Sensitivity Analysis in Observational 
   Research: Introducing the E-value. Annals of Internal Medicine.
